### 2022-08-19 main/chapter_searches.tex
add more prob words; add a sig bkg figure to explain prior likelihood difference

We begin with what probability is.
At its most abstract, a probability a proportion of stuff within a larger
whole, and probabilities are useful because they can be manipulated with an
algebra that corresponds to how one naturally manipulates such proportions.

\paragraph{Probability, abstract}
A probability $\prob{b}{a}$ is a number that quantifies the proportion
of $b$ within $a$, and can be expressed as a ratio of measures
as~\cite{axioms1010038}
\begin{equation}
\label{eqn:searches_prob_ratio}
\prob{b}{a} = \frac{m(b, a)}{m(a)},
\end{equation}
in which $m(x)$ is a measure (a non-negative quantity assigned to its
argument, such as mass of tea in a cup), and the comma ``,'' means `and'
(logical and or set intersection or any other equivalent operation).
To continue that example, $\prob{b}{a}$ could be a proportion of tea within
a spoon within a cup, where $m(b \land a)$ measures the volume of tea in a
teaspoon that is partially submerged in the cup.%
\footnote{%
All probabilities are conditional in this definition; $\probc{b}$ alone does
not exist unless the conditioning information $a$ is implicit from context.%
}

@@

Probability algebra has two rules that can derive from the measure ratio of
Equation~\ref{eqn:searches_prob_ratio}:
a \textbf{product rule}
\begin{equation}
\prob{c, b}{a} = \prob{c}{b, a}\times \prob{b}{a},
\end{equation}
and a \textbf{sum rule}
\begin{equation}
\prob{c\textrm{ or }b}{a} = \prob{c}{a} + \prob{b}{a} - \prob{c, b}{a}.
\end{equation}
To avoid of this subtraction we prefer split any problem into orthogonal or
disjoint states; for such cases, the normalization condition is that
\begin{equation}
\sum_i \prob{b_i}{a} = 1.
\end{equation}
These rules are basic and well known, but we state them here to support the
following statements of how probabilities can be applied for practical
results.%
\footnote{%
The formulation with sums and products is not strictly unique, but all
alternatives are equivalent up to an invertible
transformation~\cite{axioms1010038}.
This choice is usually best because it uses ordinary addition and
multiplication.
In numerical analysis, however, it is often practical to use $\log$
probabilities (to avoid overflows outside the range representable by floating
point numbers(, thus changing multiplication to addition, and addition to
the `$\log$-add-$\exp$' operation.
}

\paragraph{Probability, applied}
% science snake

Probabilistic data analysis proceeds as follows:
\begin{enumerate}
\item Predict new data $d$ with probability $\prob{d}{\ldots}$.
\item Update your predictions with the observed data to $\prob{x}{d, \ldots}$.
\item Go to 1.
\end{enumerate}

% evidence

>>>>

Statistical data analysis suffers from dogma --- prescriptions and recipes that
students are taught to memorize and follow without knowing their reasons or
limitations.
From a standard undergraduate physics education, my experience has been
error propagation formulae,
least squares fitting routines,
chi-squared per degree of freedom,
and myriad statistical tests and confidence intervals.
Many of these are effective approximations, but all also fail if misused
outside of their domain.%
\footnote{%
Multiplicative error propagation is an example.
For $a = \mu_a \pm \sigma_a$ and $b = \mu_b \pm \sigma_b$,
the perturbative approximation is
\(
\widetilde{\sigma}_{ab}
= \mu_a \mu_b \sqrt{\sigma_a^2/\mu_a^2 + \sigma_b^2/\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2
\),
but calculation with their moments derives
\(
\sigma_{ab}
= \sqrt{(\mu_a^2 + \sigma_a^2)(\mu_b^2 + \sigma_b^2) - \mu_a^2\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2 + \sigma_a^2 \sigma_b^2
\)
exactly if independent.
Since $\sigma_{ab}^2 = \widetilde{\sigma}_{ab}^2 + \sigma_a^2 \sigma_b^2$, the
propagation formula works only for small relative errors.
}
All effective methods of data analysis can be derived upwards from the roots of
probability and decision theory, and as elsewhere in physics, by knowing the
derivations we can also know when each approximation should work and when it
needs replacing.

We begin with what probability is: at its most abstract, a probability a
proportion of stuff within a larger whole.
Probabilities useful because proportions have applications, and because they
can be manipulated and interrelated with a linear and intuitive algebra.

\paragraph{Probability, abstract}
A probability $\prob{b}{a}$ is a number that quantifies the proportion
of $b$ within $a$, and can be expressed as a ratio~\cite{axioms1010038}
\begin{equation}
\label{eqn:searches_prob_ratio}
\prob{b}{a} = \frac{m(b, a)}{m(a)},
\end{equation}
in which $m(x)$ is a measure (a non-negative quantity assigned to its
argument, such as mass of tea in a cup), and the comma `,' means `and'
(logical and or set intersection or any other equivalent operation).
To continue that example, $\prob{b}{a}$ could be a proportion of tea within
a spoon within a cup, where $m(b, a)$ measures the volume of tea in a
teaspoon that is partially submerged in the cup.%
\footnote{%
All probabilities are conditional in this definition; $\probc{b}$ alone does
not exist unless the conditioning information $a$ is implicit from context.%
}

@@

Two algebraic rules derive from Equation~\ref{eqn:searches_prob_ratio}:
a \textbf{product rule}
\begin{align}
\prob{c, b}{a} &= \prob{c}{b, a}\times \prob{b}{a},
\label{eqn:searches_product_rule}
\intertext{%
that describes nested proportions
(perhaps a spoon within a teacup within a sink),
and a \textbf{sum rule}%
}
\prob{c\vee b}{a} &= \prob{c}{a} + \prob{b}{a} - \prob{c, b}{a},
\label{eqn:searches_sum_rule}
\end{align}
that describes combination in a union $\vee$ that coincides with logical `or'.
The subtraction avoids double-counting of any overlap, and is often avoided by
splitting any problem into orthogonal, or disjoint states.
For disjoint cases with labels $b_i$, normalization therefore requires that
\begin{equation}
\label{eqn:searches_normalization}
\sum_i \prob{b_i}{a} = 1.
\end{equation}
These rules are basic and well known, but we state them here to support the
following statements of how probabilities can be applied for practical
results.%
\footnote{%
The formulation with sums and products is not strictly unique, but all
alternatives are equivalent up to an invertible
transformation~\cite{axioms1010038}.
This choice is usually best because it uses ordinary addition and
multiplication.
In numerical analysis, however, it is often practical to use $\log$
probabilities to avoid overflows outside the range representable by floating
point numbers, thus changing multiplication to addition and addition to
the `$\log$-add-$\exp$' operation.
}

\paragraph{Probability, applied}
% science snake
Science involves making predictions, testing those predictions against data,
and reporting what is learned from the results.
Various probabilities are involved in this prediction and learning that have
conventional names to communicate what we intend to do with them.
Most important for prediction is the \textbf{likelihood}
\begin{align}
L(a) &= \prob{b}{a}
\intertext{%
that explores the probabilities assigned to a fixed result $b$ from different
contexts $a$, or how well different hypotheses predicted a result.
A likelihood is clearly a probability, but it is not a probability distribution
--- only the left hand argument $b$ forms a normalized distribution.
Alternatively, fixing the context gives a \textbf{prior}%
}
\pi(b) &= \prob{b}{a}
\end{align}
for fixed $a$, which is a normalized distribution over disjoint states $b$.

Likelihood and prior are evidently related --- they explore two indices into
the same object, two sides of the same coin, and are assigned by the same
principles.
But they are profoundly different objects that work for different purposes,
and clarity in their distinction will help in our later understanding of
practical methods.
For observed data, we use likelihoods to compare which hypotheses predicted
them better than others, and before observing data we use priors to describe
what data might result from different models to design our experiments.

Take for example an overflow histogram bin that requires $\met > 100\,\eV[G]$,
and is part of a figure with stacked histograms from background
and (supersymmetric) signal samples.
If those histograms are normalized to unit area, they are prior distributions
for events from the background and signal hypotheses.
The likelihood function for this overflow bin is
$\prob{\met > 100\,\eV[G]}{a}$, and depends on the hypothetical origin of that
event: $a = \textrm{`background'}$ or $a = \textrm{`signal'}$.
\TODO{transformation behaviours are important}

Figure~\ref{fig:searches_sig_bkg_prior_likelihood}

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figures/searches_sig_bkg_prior_likelihood.pdf}
\caption[
Background and signal prior densities
]{%
Background and signal prior densities in $\eV[G]$ and $\eV[G]^{1/2}$ units.
When changing coordinates, the probability is a fluid of constant total volume.
In each bin a likelihood function is indexed by the background--signal labels
and is independent of the prior units since the area in each bin is invariant.
}
\label{fig:searches_sig_bkg_prior_likelihood}
\end{figure}

% evidence: lower level events with different $\met$ values, bin likelihood

% likelihood ratio
\begin{equation}
\hphantom{\quad \quad \mid\mid \mathrm{context}}
\frac{\prob{S}{d}}{\prob{B}{d}}
=
\!\!
\underbrace{\frac{\prob{d}{S}}{\prob{d}{B}}}_\textrm{likelihood ratio}
\!\!\times~
\frac{\probc{S}}{\probc{B}}
\quad \quad \mid\mid \mathrm{context}
\end{equation}

Density is a misleading term. Prior mass cannot be compressed, it can only be
moved and redistributed. It is more like depth of tea than density of air.

Probabilistic learning proceeds as follows:
\begin{enumerate}
\item Predict new data $d$ with probability $\prob{d}{\ldots, h}$.
\item Update your predictions with the observed data to $\prob{x}{d, \ldots, h}$.
\item Go to 1.
\end{enumerate}
